---
title: "Final Project EDA"
author: "Aakash Ahuja & Ayush Bhatia"
date: "18/04/2020"
output: html_document
---

#Getting the Data
```{r,message=F, warning=F}
library(readxl)
library(dplyr)
library(ggplot2)
#install.packages('imputeTS')
#install.packages('ggpmisc')
#install.packages('rlang')
library(imputeTS)
full_data<-readxl::read_xlsx('EDA Final Project Data.xlsx', sheet = 2)
Productivity<-readxl::read_xlsx('EDA Final Project Data.xlsx', sheet = 4)
colnames(Productivity)[1]<-"State Name"
colnames(full_data)[1]<-"Year"
colnames(Productivity)[7]<-"Productivity_Value"
full_data$`Measure Name`<-factor(full_data$`Measure Name`)

Productivity<-filter(Productivity,`State Name`!='District of Columbia' &  `Units`=='2012=100')

full_data<-full_data[,c(1,3,4,5,6,7)]

Combined_Productivity<-left_join(Productivity,full_data, by= c("State Name","Year"))
```

**Definition of Air Pollution Value:** Average exposure of the general public to particulate matter of 2.5 microns or less (PM2.5) measured in micrograms per cubic meter (3-year estimate)

**Definition of Binge Drinking Value:** Percentage of adults who reported either binge drinking (having four or more [women] or five or more [men] drinks on one occasion in the past 30 days)

**Definition of Diabetes Value:** Percentage of adults who reported being told by a health professional that they have diabetes (excluding prediabetes and gestational diabetes)

**Definition of High Blood Pressure Value:** Percentage of adults who reported being told by a health professional that they have High Blood Pressure

**Definition of High Cholestrol Value:** Percentage of adults who reported being told by a health professional that they have High Cholestrol

**Definition of Obesity Value:** Percentage of adults with a body mass index of 30.0 or higher based on reported height and weight

**Definition of Physical Inactivity:** Percentage of adults who reported doing no physical activity or exercise other than their regular job in the past 30 days

**Definition of Smoking Value:** Percentage of adults who are smokers (reported smoking at least 100 cigarettes in their lifetime and currently smoke daily or some days)

**Definition of Smoking Value:** Percentage of adults who are smokers (reported smoking at least 100 cigarettes in their lifetime and currently smoke daily or some days)

**Definition of Violent Crime value:** Number of murders, rapes, robberies and aggravated assaults per 100,000 population

**Definition of Frequent Mental distress value:**Percentage of adults who reported their mental health was not good 14 or more days in the past 30 days

**Definition of Insufficient Sleep value:**Percentage of adults who reported sleeping less than seven hours in a 24-hour period on average

**Definition of Water Fluoridation value:**Percentage of population served by community water systems that receive fluoridated water


```{r,message=F, warning=F}
library(scales)
US_Mean_Productivity<-aggregate(as.numeric(Productivity_Value)~Year,FUN = mean, data = Productivity)
US_Mean_Productivity

US_Mean_Productivity$Year<- as.Date(as.character(US_Mean_Productivity$Year), format = "%Y")

ggplot(US_Mean_Productivity,aes(x = as.Date(Year), y =`as.numeric(Productivity_Value)`))+
  geom_line(color='red')+
  scale_x_date(labels = date_format("%y"))+
  xlab('Year')+
  ylab('Productivity')+
  labs(title = "US Productivity across 2007-2017")

```

**The overall trend for US Productivity is that the productivity for the country is increasing across 10 years.**


**Making a plotting function to check this trend for all factors in our dataset faceted by year**
```{r,message=F, warning=F}
library(plotly)

#Making a plotting function for each of the measures like Air Pollution, Water Quality etc
plottng_function<-function(Name){
  
Measure_df<-Combined_Productivity[Combined_Productivity$`Measure Name`==as.character(Name) & Combined_Productivity$Year!=2012,]

p<-ggplot(Measure_df, aes(x = as.numeric(Value), y = as.numeric(Productivity_Value)))+
  geom_point(size=0.5, aes(color=`State Name`, text=paste("Productivity:", round(as.numeric(Productivity_Value),1),"<br />", "Value",Value,"<br />")), show.legend = F)+
  geom_smooth(method = "lm", se=F, linetype='solid')+
  facet_wrap(~Year, ncol = 5)+
  ylab('Productivity Value')+
  xlab(paste(Name,"Value"))+
  theme(legend.position = 'None')
gp<-hide_legend(ggplotly(p,tooltip=c('text', 'State Name') ))
#gppp<-style( gpp, text=mytext, hoverinfo = "text")
print(gp)
}

```

**Making a list of all the factors & plotting it against productivity value faceted by year**

```{r,fig.width=15, fig.height=8, echo=FALSE,message=F, warning=F}
my_list<-c("Air Pollution","Binge Drinking","Diabetes","High Blood Pressure","High Cholesterol","Obesity","Physical Inactivity","Smoking","Violent Crime","Frequent Mental Distress","Insufficient Sleep","Water Fluoridation")

plotlist_1<-list()

for (i in my_list) {
  Measure_df<-Combined_Productivity[Combined_Productivity$`Measure Name`==as.character(i) & Combined_Productivity$Year!=2012,]
  p<-ggplot(Measure_df, aes(x = as.numeric(Value), y = as.numeric(Productivity_Value)))+
  geom_point(size=0.5, aes(color=`State Name`, text=paste("Productivity:", round(as.numeric(Productivity_Value),1),"<br />", "Value",Value,"<br />")), show.legend = F)+
  geom_smooth(method = "lm", se=F, linetype='solid')+
  facet_wrap(~Year, ncol = 5)+
  ylab('Productivity Value')+
  xlab(paste(i,"Value"))+
  theme(legend.position = 'None')
plotlist_1[[i]]<-hide_legend(ggplotly(p,tooltip=c('text', 'State Name') ))
print(plotlist_1[[i]]) 
}

htmltools::tagList(setNames(plotlist_1, NULL))

```

**Based on these graphs, we see some clear & obvious trends & would like to remove certain variables from consideration based on lack of trend.**

**The General trend of productivity across the last 3 years is that the mean productivity for all states is increasing with time as can be seen from the first plot. So if for a feature the productivity decreases, there could definitely be a relation.**

**Features we would like to consider for further analysis**
1. Air Pollution: Yes
2. Binge Drinking: Yes
3. Diabetes: No: No clear trend shown across the years
4. High Blood Pressure: No clear trend shown across the years
5. High Cholestrol Value: Yes but the trend seems to be mixed
6. Obesity: Yes
7. Physical Inactivity: Yes
8. Smoking: Yes
9. Violent Crime: Yes
10. Frequent Mental Distress: Yes
11. Insufficient Sleep: Yes
12. Water Fluoridation: No. Lack of data even though the trend seems interesting


**Percentage Change**
```{r,message=F, warning=F}
Combined_Productivity <- na.omit(Combined_Productivity)

remove_measures <- c("Diabetes","High Blood Pressure","Excessive Drinking","Water Fluoridation","Poor Mental Health Days")

`%notin%` <- Negate(`%in%`)

prod_data_perc <- Combined_Productivity %>% filter(
  `Measure Name` %notin% remove_measures
  )

Measure_names<-as.vector(unique(prod_data_perc$`Measure Name`))

#measure_names <- measure_max$`Measure Name`
#state_names <- measure_max$`State Name`
#measure_change <- measure_max$Value - measure_min$Value
#measure_perc_change <- (measure_change*100)/measure_min$Value

#prod_change <- as.numeric(measure_max$Productivity_Value) - as.numeric(measure_min$Productivity_Value)

#prod_perc_change <- (prod_change*100)/as.numeric(measure_min$Productivity_Value)

#final_perc_data <- data.frame("Measure Name" = measure_names,"State Name" = state_names,"Productivity Change" = prod_perc_change, "Measure Change" = measure_perc_change )

```

**Plotting Percentage change in productivity and measure values**



```{r,fig.width=15, fig.height=8,message=F, warning=F}
library(ggpmisc)

average_change_by_measure<-aggregate(cbind(as.numeric(Productivity_Value), as.numeric(Value) ) ~ `State Name`+`Measure Name` , data = prod_data_perc, FUN =  mean)
colnames(average_change_by_measure)[3]<-"Mean_Productivity"
colnames(average_change_by_measure)[4]<-"Mean_Values"


plotlist_3 = list()
for (i in Measure_names) {
  
  temp_df_mean<-average_change_by_measure[average_change_by_measure$`Measure Name`==i,]
  p<-ggplot(temp_df_mean, aes(x=Mean_Values, y=Mean_Productivity)) +
    stat_quadrant_counts(geom='text', colour = "blue", size=5, xintercept = mean(temp_df_mean$Mean_Values), yintercept = mean(temp_df_mean$Mean_Productivity)) +
  geom_point(size=0.75, aes(color=`State Name`)) +
  lims(x=c(min(temp_df_mean$Mean_Values)-0,max(temp_df_mean$Mean_Values)+0),y=c(min(temp_df_mean$Mean_Productivity)-0,max(temp_df_mean$Mean_Productivity)+0)) +
  xlab(paste(i,"Mean"))+  
  theme_minimal() +
  #coord_fixed() +  
  theme(legend.position = 'None')+
  geom_vline(xintercept = mean(temp_df_mean$Mean_Values)) + geom_hline(yintercept = mean(temp_df_mean$Mean_Productivity) )
 plotlist_3[[i]]<- hide_legend(ggplotly(p,tooltip=c('State Name') ))
 print(plotlist_3[[i]]) 
 #gp<-hide_legend(ggplotly(p,tooltip=c('State Name') ))
  #print(gp)
}
htmltools::tagList(setNames(plotlist_3, NULL))

```

**Finally we come to building a final model for productivity values of different states of United States**
```{r,message=F, warning=F}
library(imputeTS)
library(broom)
library(MASS)
library(tidyverse)
library(GGally)
model_data<-prod_data_perc[,c("State Name","Year","Measure Name","Productivity_Value","Value")]
model_data$Productivity_Value<-as.numeric(model_data$Productivity_Value)

selected_model_data <- model_data %>%
  pivot_wider(names_from = `Measure Name`, values_from = Value)

selected_model_data<-na_mean(selected_model_data)
model_lm_data_1<-selected_model_data[,-c(1,2)]
model_lm_data<-selected_model_data
#Checking Correlation between variables
ggpairs(model_lm_data_1)
#Air Pollution Obesity, Smoking, Violent Crime, Insufficient Sleep

#Fitting a model
mod_fit<-lm(Productivity_Value~., data = model_lm_data)
#summary(mod_fit)

final_fit<-augment(mod_fit)
```

```{r,message=F, warning=F}

ggplot(final_fit, aes(x = Air.Pollution, y = .resid)) + geom_point(size =1.5)  + geom_smooth(method ="rlm")+labs(title="Air Pollution Vs Residuals")

ggplot(final_fit, aes(x = Obesity, y = .resid)) + geom_point(size =1.5)  + geom_smooth(method ="rlm")+labs(title="Obesity Vs Residuals")

ggplot(final_fit, aes(x = Smoking, y = .resid)) + geom_point(size =1.5)  + geom_smooth(method ="rlm", se=F)+labs(title="Smoking Vs Residuals")

ggplot(final_fit, aes(x = Violent.Crime, y = .resid)) + geom_point(size =1.5)  + geom_smooth(method ="rlm")+labs(title="Violent Crime Vs Residuals")

ggplot(final_fit, aes(x = Insufficient.Sleep, y = .resid)) + geom_point(size =1.5)  + geom_smooth(method ="rlm", se=F)+labs(title="Insufficient Sleep Vs Residuals")
```
